Stability of Ensembles of Biological Networks

### 2014-09-11 journal.tex
modify title
fix edge case where zero eigenvalues are slightly negative due to numerical error
fix off-by-one error in connectivity value of latex table
add plot of stability to perturbations for 3x3 systems
mention case of three component systems in abstract

\title{Stability of Dynamical System Ensembles}

@@

\begin{abstract}
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having many parameters and thus many degrees of freedom. When the details of a system are unknown, one effective approach is to study the dynamical properties of a collection of models determined by constraints applying to all such systems. Here we analyze the stability of a large class of dynamical systems to perturbations in the underlying structure of the system: a property referred to as \emph{structural stability}. In particular, we precisely determine the probability distribution over system connectivity, a parameter which has significant implications from the study of gene-regulatory networks to ecosystem dynamics, of structural stability for systems with two interacting components. We show in this case that structural stability has a non-monotonic relationship with system connectivity. These results support future work attempting to characterize the scaling relationship between structural stability and system connectivity. The latter investigation is necessary to evaluate several conjectured but ultimately untested hypotheses about biological networks.
\end{abstract}

@@

The probability of \emph{a priori} stability and of stability to perturbations for all three component systems is given in Table \ref{tab:structstabmat3}.

% \begin{table}[h]
\begin{longtable}{ c || c | c | c }
\hline
matrix & connectivity & \specialcell{probability of stability\\to perturbation} & probability of stability\\
\hline
$\begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 0\\
0 & 0 & 1\\
\end{pmatrix}$ & 4 & 0.496 & 0.125\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 0 & 1\\
1 & 1 & 0\\
\end{pmatrix}$ & 5 & 0.117 & 0.07\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 0\\
1 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.49 & 0.123\\
$\begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 1\\
0 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.62 & 0.128\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 0 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.32 & 0.089\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.599 & 0.123\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 1\\
1 & 1 & 0\\
\end{pmatrix}$ & 6 & 0.499 & 0.127\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 0 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.398 & 0.064\\
$\begin{pmatrix}
0 & 1 & 0\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.431 & 0.08\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 0\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.597 & 0.126\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 0 & 0\\
\end{pmatrix}$ & 6 & 0.41 & 0.061\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 1\\
0 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.7 & 0.122\\
$\begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 1\\
0 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.595 & 0.123\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 0\\
0 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.701 & 0.125\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.574 & 0.117\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 0 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.486 & 0.075\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 7 & 0.539 & 0.092\\
$\begin{pmatrix}
0 & 1 & 0\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.497 & 0.088\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 0\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.664 & 0.126\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 1 & 0\\
\end{pmatrix}$ & 7 & 0.482 & 0.086\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 1 & 0\\
1 & 0 & 1\\
\end{pmatrix}$ & 7 & 0.589 & 0.126\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 1\\
0 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.66 & 0.127\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
0 & 0 & 1\\
\end{pmatrix}$ & 7 & 0.749 & 0.121\\
$\begin{pmatrix}
1 & 1 & 0\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 7 & 0.61 & 0.097\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 8 & 0.571 & 0.102\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 0 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 8 & 0.472 & 0.077\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 8 & 0.59 & 0.11\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
0 & 1 & 1\\
\end{pmatrix}$ & 8 & 0.709 & 0.125\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 8 & 0.642 & 0.122\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 8 & 0.63 & 0.108\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 9 & 0.57 & 0.092\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 9 & 0.649 & 0.111\\
$\begin{pmatrix}
1 & 1 & 1\\
1 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 10 & 0.645 & 0.103\\
\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for three component systems estimated via Monte Carlo sampling.} All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat3}
\end{longtable}
%
% \end{table}

>>>>

\title{Stability of Ensembles of Biological Networks}

@@

\begin{abstract}
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having many parameters and thus many degrees of freedom. When the details of a system are unknown, one effective approach is to study the dynamical properties of a collection of models determined by constraints applying to all such systems. Here we analyze the stability of a large class of dynamical systems to perturbations in the underlying structure of the system: a property referred to as \emph{structural stability}. In particular, we precisely determine the probability distribution over system connectivity, a parameter which has significant implications from the study of gene-regulatory networks to ecosystem dynamics, of structural stability for systems with two and three interacting components. We show for these cases that structural stability has a non-monotonic relationship with system connectivity. These results support future work attempting to characterize the scaling relationship between structural stability and system connectivity. The latter investigation is necessary to evaluate several conjectured but ultimately untested hypotheses about biological networks.
\end{abstract}

@@

The probability of \emph{a priori} stability and of stability to perturbations for all three component systems is given in Table \ref{tab:structstabmat3}. Fig. \ref{fig:stab3x3} shows the probability of stability to perturbations for all three component systems.

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.8\columnwidth]{fig/stab3x3.pdf}
\caption{{\bf Plot of stability to perturbations versus connectivity for three component systems.} }
\label{fig:stab3x3}
\end{figure}

% \begin{table}[h]
\begin{longtable}{ c || c | c | c }
\hline
matrix & connectivity & \specialcell{probability of stability\\to perturbation} & probability of stability\\
\hline
$\begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 0\\
0 & 0 & 1\\
\end{pmatrix}$ & 3 & 0.499 & 0.126\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 0\\
1 & 0 & 1\\
\end{pmatrix}$ & 4 & 0.505 & 0.121\\
$\begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 1\\
0 & 0 & 1\\
\end{pmatrix}$ & 4 & 0.622 & 0.127\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.595 & 0.121\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 1\\
1 & 1 & 0\\
\end{pmatrix}$ & 5 & 0.494 & 0.128\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 0 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.41 & 0.061\\
$\begin{pmatrix}
0 & 1 & 0\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.43 & 0.078\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 0\\
1 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.605 & 0.12\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 0 & 0\\
\end{pmatrix}$ & 5 & 0.405 & 0.06\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 1\\
0 & 0 & 1\\
\end{pmatrix}$ & 5 & 0.698 & 0.122\\
$\begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 1\\
0 & 1 & 1\\
\end{pmatrix}$ & 5 & 0.587 & 0.128\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 0\\
0 & 1 & 1\\
\end{pmatrix}$ & 5 & 0.707 & 0.127\\
$\begin{pmatrix}
0 & 0 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.578 & 0.121\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 0 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.487 & 0.081\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.543 & 0.098\\
$\begin{pmatrix}
0 & 1 & 0\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.501 & 0.088\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 0\\
1 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.662 & 0.123\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 1 & 0\\
\end{pmatrix}$ & 6 & 0.467 & 0.079\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 1 & 0\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.583 & 0.13\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 1\\
0 & 1 & 1\\
\end{pmatrix}$ & 6 & 0.659 & 0.124\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
0 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.751 & 0.124\\
$\begin{pmatrix}
1 & 1 & 0\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 6 & 0.604 & 0.097\\
$\begin{pmatrix}
0 & 1 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.564 & 0.103\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 0 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.475 & 0.068\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 7 & 0.591 & 0.108\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
0 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.717 & 0.119\\
$\begin{pmatrix}
1 & 0 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 7 & 0.648 & 0.122\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
1 & 0 & 1\\
\end{pmatrix}$ & 7 & 0.627 & 0.105\\
$\begin{pmatrix}
0 & 1 & 1\\
1 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 8 & 0.577 & 0.093\\
$\begin{pmatrix}
1 & 1 & 1\\
0 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 8 & 0.639 & 0.109\\
$\begin{pmatrix}
1 & 1 & 1\\
1 & 1 & 1\\
1 & 1 & 1\\
\end{pmatrix}$ & 9 & 0.638 & 0.106\\
\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for three component systems estimated via Monte Carlo sampling.} All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat3}
\end{longtable}
%
% \end{table}
